Short-time simulations of open collective quantum spin systems

### 2019-04-23 soc_squeezing/collective_paper/collective_spin.tex
minor corrections and wording changes

The time derivative operator $\check T$ will generally couple spin
operators $\hat\S_{\v n}$ to spin operators $\hat\S_{\v m}$ with
higher ``weight'', i.e.~with $\abs{\v m}>\abs{\v n}$, where
$\abs{\v\ell}\equiv\sum_\a\ell_\a$.  The growth of operator weight
signifies the growth of many-body correlations.  In practice, keeping
track of this growth will eventually require more computational
resources than are available, meaning we must somehow truncate our
equations of motion.  The simplest truncation strategy would be to
take
\begin{align}
  \f{d}{dt} \hat\O\p{t}
  \to \sum_{w\p{\v m}<W} \hat\S_{\v m}
  \sum_{\v n} T_{\v m\v n} \hat\O_{\v n}\p{t}
  \label{eq:weight_truncation}
\end{align}
for some weight measure $w$, e.g.~$w\p{\v m}=\abs{\v m}$, and a
high-weight cutoff $W$.  The truncation in
Eq.~\eqref{eq:weight_truncation} closes the system of differential
equations defined by Eq.~\eqref{eq:EOM}, and allows us to solve it
using standard numerical methods.  Some initial conditions for this
system of differential equations, namely expectation values of
collective spin operators with respect to spin-polarized (Gaussian)
states that are generally simple to prepare experimentally, are
provided in Appendix \ref{sec:initial_conditions}.

@@

\begin{figure*}
  \centering
  \subfloat[Two-time correlator
  $C\p{t}\equiv\abs{C\p{t}}\exp\sp{i\phi\p{t}}$ defined in
  Eq.~\eqref{eq:two_time}. \label{fig:two_time}]
  {\includegraphics{two_time.pdf}} \\
  \subfloat[Squared commutator $D\p{t}$ defined in
  Eq.~\eqref{eq:OTOC}. \label{fig:OTOC}]
  {\includegraphics{OTOC.pdf}}
  \caption{The two-time connected correlator $C\p{t}$ and squared
    commutator $D\p{t}$, respectively defined in
    Eqs.~\eqref{eq:two_time} and \eqref{eq:OTOC}, for $N=10^4$ spins
    initially in the polarized state
    $\ket{\X}\propto\p{\ket\up+\ket\dn}^{\otimes N}$ evolving under
    the squeezing Hamiltonians in Eqs.~\eqref{eq:OAT}--\eqref{eq:TNT}.
    Results are shown for time evolution without decoherence (leftmost
    panels), as well as with decoherence at rates $\Gamma_\a$ or
    $\gamma_\a$, respectively corresponding to collective or
    single-spin decoherence through jump operators $\hat S_\a$ or
    $\set{\hat s_\a^{(j)}}$.  Collective decoherence rates
    ($\Gamma_\a$) were chosen to be comparable to the OAT squeezing
    strength $\chi$, while single-spin decoherence rates ($\gamma_\a$)
    were scaled by an additional factor of $S=N/2$ in order to yield
    comparable effects on the correlators $C\p{t}$ and $D\p{t}$ at the
    times shown.  At short times, the multi-time correlations captured
    by $C\p{t}$ and $D\p{t}$ are essentially insensitive to collective
    dephasing ($\Gamma_\z$), while their magnitudes are generally
    suppressed by other forms of decoherence.  Whether decoherence
    occurs through spin decay, excitation, or dephasing also leaves a
    signature on the phase $\phi\p{t}$ of $C\p{t}$.  Results computed
    using the TST expansion in Eq.~\eqref{eq:TST} with $M=20$.}
  \label{fig:multi_time}
\end{figure*}

As a final example of collective-spin physics that is numerically
accessible via the TST expansion of Heisenberg operators, we consider
the calculation of two-time correlation functions and
out-of-time-ordered correlators (OTOCs).  In particular, we consider
the effect of decoherence on short-time behavior of the two-time
connected correlator
\begin{align}
  C\p{t}
  \equiv \f1S\p{\bk{\hat S_+\p{t} \hat S_-\p{0}}
    - \bk{\hat S_+\p{t}}\bk{\hat S_-\p{0}}},
  \label{eq:two_time}
\end{align}
and the expectation value of a squared commutator,
\begin{align}
  D\p{t}
  \equiv \f1{S^2}\Braket{\sp{\hat S_+\p{t},\hat S_-\p{0}}_-^\dag
    \sp{\hat S_+\p{t},\hat S_-\p{0}}_-},
  \label{eq:OTOC}
\end{align}
in the context of the squeezing models in Section \ref{sec:squeezing}.
In an equilibrium setting, correlation functions similar to that in
Eq.~\eqref{eq:two_time} contain information about the linear response
of Heisenberg operators to perturbations of a system; in a
non-equilibrium setting, they contribute to short-time linear response
(see Appendix \ref{sec:linear_response}).  Similar correlators have
made appearances as order parameters that diagnose time-crystalline
phases of matter\cite{tucker2018shattered}.  OTOCs such as that in
Eq.~\eqref{eq:OTOC}, meanwhile, are commonly examined for signatures
of quantum chaos and information scrambling\cite{maldacena2016bound,
  swingle2018unscrambling, garcia-mata2018chaos}.

@@

In order to compute a quantity such as spin squeezing, we need to
compute correlators of the form $\bk{\S_{\v n}\p{t}}$, where
$\S_{\v n}\p{t}$ is the time-evolved Heisenberg operator corresponding
to $\S_{\v n}$, with $\S_{\v n}\p{0}\equiv\S_{\v n}$.  The order-$M$
truncated short-time (TST) expansion takes
\begin{align}
  \bk{\S_{\v n}\p{t}}
  \to \sum_{k=0}^M \f{t^k}{k!} \sum_{\v m} \bk{\S_{\v m}} T^k_{\v m\v n},
  \label{eq:tutorial_TST}
\end{align}
where $T^k_{\v m\v n}$ are matrix elements of the $k$-th time
derivative operator, given by
\begin{align}
  T^0_{\v m\v n} \equiv
    \begin{cases}
      1 & \v m = \v n, \\
      0 & \t{otherwise}
    \end{cases},
  &&
  T^1_{\v m\v n} \equiv T_{\v m\v n},
  &&
  T^{k>1}_{\v m\v n}
  \equiv \sum_{\v p_1,\v p_2,\cdots,\v p_{k-1}}
  T_{\v m\v p_{k-1}} \cdots T_{\v p_3\v p_2}
  T_{\v p_2\v p_1} T_{\v p_1\v n}.
\end{align}
Matrix elements $T^k_{\v m\v n}$ and initial-time expectation values
$\bk{\S_{\v m}}$ are thus computed on an as-needed basis for any
particular correlator $\bk{\S_{\v n}\p{t}}$ of interest, and combined
according to \eqref{eq:tutorial_TST}.  Note that initial-time
expectation values $\bk{\S_{\v m}}$ are an input to the TST expansion,
and need to be computed separately for any given initial state;
expectation values with respect to spin-polarized (Gaussian) states
are provided in Appendix \ref{sec:initial_conditions}.  In practice,
we further collect terms in \eqref{eq:tutorial_TST} to write
\begin{align}
  \bk{\S_{\v n}\p{t}} \to \sum_{k=0}^M c_{\v n k} t^k,
  &&
  c_{\v n k} \equiv \f1{k!} \sum_{\v m} \bk{\S_{\v m}} T^k_{\v m\v n},
\end{align}
where $c_{\v n k}$ are time-independent coefficients for the
polynomial time expansion of $\bk{\S_{\v n}\p{t}}$.  After computing
the coefficients $c_{\v n k}$, there is only negligible computational
overhead to compute the correlator $\bk{\S_{\v n}\p{t}}$ for any time
$t$ up to the breakdown time $t_M$ of an order-$M$ TST expansion.

>>>>

The time derivative operator $\check T$ will generally couple spin
operators $\hat\S_{\v n}$ to spin operators $\hat\S_{\v m}$ with
higher ``weight'', i.e.~with $\abs{\v m}>\abs{\v n}$, where
$\abs{\v\ell}\equiv\sum_\a\ell_\a$.  The growth of operator weight
signifies the growth of many-body correlations.  In practice, keeping
track of this growth will eventually require more computational
resources than are available, meaning we must somehow truncate our
equations of motion.  The simplest truncation strategy would be to
take
\begin{align}
  \f{d}{dt} \hat\O\p{t}
  \to \sum_{w\p{\v m}<W} \hat\S_{\v m}
  \sum_{\v n} T_{\v m\v n} \O_{\v n}\p{t}
  \label{eq:weight_truncation}
\end{align}
for some weight measure $w$, e.g.~$w\p{\v m}=\abs{\v m}$, and a
high-weight cutoff $W$.  The truncation in
Eq.~\eqref{eq:weight_truncation} closes the system of differential
equations defined by Eq.~\eqref{eq:EOM}, and allows us to solve it
using standard numerical methods.  Some initial conditions for this
system of differential equations, namely expectation values of
collective spin operators with respect to spin-polarized (Gaussian)
states that are generally simple to prepare experimentally, are
provided in Appendix \ref{sec:initial_conditions}.

@@

\begin{figure*}
  \centering
  \subfloat[Two-time correlator
  $C\p{t}\equiv\abs{C\p{t}}\exp\sp{i\phi\p{t}}$ defined in
  Eq.~\eqref{eq:two_time}. \label{fig:two_time}]
  {\includegraphics{two_time.pdf}} \\
  \subfloat[Squared commutator $D\p{t}$ defined in
  Eq.~\eqref{eq:four_point}. \label{fig:four_point}]
  {\includegraphics{OTOC.pdf}}
  \caption{The two-time connected correlator $C\p{t}$ and squared
    commutator $D\p{t}$, respectively defined in
    Eqs.~\eqref{eq:two_time} and \eqref{eq:four_point}, for $N=10^4$
    spins initially in the polarized state
    $\ket{\X}\propto\p{\ket\up+\ket\dn}^{\otimes N}$ evolving under
    the squeezing Hamiltonians in Eqs.~\eqref{eq:OAT}--\eqref{eq:TNT}.
    Results are shown for time evolution without decoherence (leftmost
    panels), as well as with decoherence at rates $\Gamma_\a$ or
    $\gamma_\a$, respectively corresponding to collective or
    single-spin decoherence through jump operators $\hat S_\a$ or
    $\set{\hat s_\a^{(j)}}$.  Collective decoherence rates
    ($\Gamma_\a$) were chosen to be comparable to the OAT squeezing
    strength $\chi$, while single-spin decoherence rates ($\gamma_\a$)
    were scaled by an additional factor of $S=N/2$ in order to yield
    comparable effects on the correlators $C\p{t}$ and $D\p{t}$ at the
    times shown.  At short times, the multi-time correlations captured
    by $C\p{t}$ and $D\p{t}$ are essentially insensitive to collective
    dephasing ($\Gamma_\z$), while their magnitudes are generally
    suppressed by other forms of decoherence.  Whether decoherence
    occurs through spin decay, excitation, or dephasing also leaves a
    signature on the phase $\phi\p{t}$ of $C\p{t}$.  Results computed
    using the TST expansion in Eq.~\eqref{eq:TST} with $M=20$.}
  \label{fig:multi_time}
\end{figure*}

% TODO: in contrast to typical OTOCs, here the operators "almost commute" -- up to 1/N terms

% TODO: mention short-time expansion, see only polynomial growth

% TODO: figures 3a, 3b: 3 panels each; no decoherence, decoherence in fig. 2, and strong decoherence

% TODO: add vertical axis labels for G_- and g_- (?)

As a final example of collective-spin physics that is numerically
accessible via the TST expansion of Heisenberg operators, we consider
the calculation of two-time correlation functions and
out-of-time-ordered correlators (OTOCs).  In particular, we consider
the effect of decoherence on short-time behavior of the two-time
connected correlator
\begin{align}
  C\p{t}
  \equiv \f1S\p{\bk{\hat S_+\p{t} \hat S_-\p{0}}
    - \bk{\hat S_+\p{t}}\bk{\hat S_-\p{0}}},
  \label{eq:two_time}
\end{align}
and the expectation value of a squared commutator,
\begin{align}
  D\p{t}
  \equiv \f1{S^2}\Braket{\sp{\hat S_+\p{t},\hat S_-\p{0}}_-^\dag
    \sp{\hat S_+\p{t},\hat S_-\p{0}}_-},
  \label{eq:four_point}
\end{align}
in the context of the squeezing models in Section \ref{sec:squeezing}.
In an equilibrium setting, correlation functions similar to that in
Eq.~\eqref{eq:two_time} contain information about the linear response
of Heisenberg operators to perturbations of a system; in a
non-equilibrium setting, they contribute to short-time linear response
(see Appendix \ref{sec:linear_response}).  Similar correlators have
made appearances as order parameters that diagnose time-crystalline
phases of matter\cite{tucker2018shattered}.  Squared commutators such
as that in Eq.~\eqref{eq:four_point}, meanwhile, are commonly examined
for signatures of quantum chaos and information
scrambling\cite{maldacena2016bound, swingle2018unscrambling,
  garcia-mata2018chaos}.

@@

In order to compute a quantity such as spin squeezing, we need to
compute correlators of the form $\bk{\S_{\v n}\p{t}}$, where
$\S_{\v n}\p{t}$ is the time-evolved Heisenberg operator corresponding
to $\S_{\v n}$, with $\S_{\v n}\p{0}\equiv\S_{\v n}$.  The order-$M$
truncated short-time (TST) expansion takes
\begin{align}
  \bk{\S_{\v n}\p{t}}
  \to \sum_{k=0}^M \f{t^k}{k!}
  \sum_{\v m} \bk{\S_{\v m}\p{0}} T^k_{\v m\v n},
  \label{eq:tutorial_TST}
\end{align}
where $T^k_{\v m\v n}$ are matrix elements of the $k$-th time
derivative operator, given by
\begin{align}
  T^0_{\v m\v n} \equiv
    \begin{cases}
      1 & \v m = \v n, \\
      0 & \t{otherwise}
    \end{cases},
  &&
  T^1_{\v m\v n} \equiv T_{\v m\v n},
  &&
  T^{k>1}_{\v m\v n}
  \equiv \sum_{\v p_1,\v p_2,\cdots,\v p_{k-1}}
  T_{\v m\v p_{k-1}} \cdots T_{\v p_3\v p_2}
  T_{\v p_2\v p_1} T_{\v p_1\v n}.
\end{align}
Matrix elements $T^k_{\v m\v n}$ and initial-time expectation values
$\bk{\S_{\v m}\p{0}}$ are thus computed as needed for any particular
correlator $\bk{\S_{\v n}\p{t}}$ of interest, and combined according
to \eqref{eq:tutorial_TST}.  Note that initial-time expectation values
$\bk{\S_{\v m}\p{0}}$ are an input to the TST expansion, and need to
be computed separately for any given initial state; expectation values
with respect to spin-polarized (Gaussian) states are provided in
Appendix \ref{sec:initial_conditions}.  In practice, we further
collect terms in \eqref{eq:tutorial_TST} to write
\begin{align}
  \bk{\S_{\v n}\p{t}} \to \sum_{k=0}^M c_{\v n k} t^k,
  &&
  c_{\v n k}
  \equiv \f1{k!} \sum_{\v m} \bk{\S_{\v m}\p{0}} T^k_{\v m\v n},
\end{align}
where $c_{\v n k}$ are time-independent coefficients for the
polynomial time expansion of $\bk{\S_{\v n}\p{t}}$.  After computing
the coefficients $c_{\v n k}$, there is only negligible computational
overhead to compute the correlator $\bk{\S_{\v n}\p{t}}$ for any time
$t$ up to the breakdown time $t_M$ of an order-$M$ TST expansion.
